\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\title{Analysis and Design of Algorithms}
\author{}
\date{April 2019}

\usepackage{natbib}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{amsmath}

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\makeatletter
\def\BState{\State\hskip-\ALG@thistlm}
\makeatother

\usepackage{listings}
\usepackage{color}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  language=Java,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

\begin{document}

\maketitle

\section{Warm up}

Lets modify the classic merge sort algorithm a little bit. What happens if instead of splitting the array in 2 parts we divide it in 3? You can assume that exists a three-way merge subroutine. What is the overall asymptotic running time of this algorithm?\\

In this version of merge sort, the recursive step of the algorithm would be called 3 times and then the sub-arrays would be merged to form the sorted original array. With this in mind, the runtime of merge sort can be calculated as follows:

\[T(n) = 3 T(n / 3) +  \Theta(n)\]

where $3 T(n / 3)$ represents the call to merge sort as the recursive step and $ \Theta(n)$ the merging of the three sorted sub-arrays. By expanding the recursive step in the definition of the runtime function, we obtain the following result:

\begin{align*}
T(n) &= 3 T(n / 3) +  \Theta(n)\\
&= 3 (3 T(n / 9) + n / 3) +  \Theta(n)\\
&= 9 T(n / 9) + 2  \Theta(n)\\
&= 27 T(n / 27) + 3  \Theta(n)
\end{align*}

which shows how each level cost is equivalent to $ \Theta(n)$. As the recursive division of three sub-arrays from the original array leads to a maximum level of $\log_3 n$, we can conclude that the asymptotic running time of three-way merge sort is $\Theta (n\log_3 n)$.

\emph{BONUS:} Implement the three-way merge sort algorithm.

\section{Competitive programming}

Welcome to your first competitive programming problem!!! 

\begin{itemize}
    \item Sign-up in Uva Online Judge (\url{https://uva.onlinejudge.org}) and in CodeChef if you want (we will use it later).
    \item Rest easy! This is not a contest, it is just an introductory problem. Your first problem is located in the ``Problems Section'' and is \textbf{100 - The 3n + 1 problem}.
    
    \begin{lstlisting}
#include <iostream>

int cycleLength(int n) {
	if (n == 1) {
		return 1;
	}
	
	return 1 + cycleLength(n % 2 ? 3 * n + 1 : n / 2);
}

using namespace std;
	
int main (void) {
	int i, j;

	while (cin >> i) {
		cin >> j;
		
		int iMin, iMax;
		
		if (i < j) {
			iMin = i;
			iMax = j;
		} else {
			iMin = j;
			iMax = i;
		}

		int max = 0;

		for (int k = iMin; k <= iMax; ++k) {
			int current = cycleLength(k);

			if (current > max) {
				max = current;
			}
		}

		cout << i << ' '<< j << ' ' << max << endl;
	}

	return 0;
}
    }
}

\end{lstlisting}
    
    \item Once that you finish with that problem continue with \textbf{458 - The Decoder}. Again, this problem is just to build your confidence in competitive programming.
    
    \begin{lstlisting}
#include <iostream>

using namespace std;

int main(void) {
	char c = cin.get();

	while (!cin.eof()) {
		if (c != '\n') {
			cout << (char) (c - 7);
		} else {
			cout << endl;
		}

		c = cin.get();
	}

	return 0;
}

\end{lstlisting}
    
    \item \emph{BONUS:} \textbf{10855 - Rotated squares}
    
\end{itemize}

\section{Simulation}

Write a program to find the minimum input size for which the merge sort algorithm always beats the insertion sort.

\begin{itemize}
    \item Implement the insertion sort algorithm
    
    \begin{lstlisting}
#include <vector>

#include "insertionSort.h"

void insertionSort(std::vector<int>& v) {
	// Current index of element to be sorted
	int curIndex;

	for (curIndex = 1; curIndex < v.size(); ++curIndex) {
		// Target index in which v[curIndex] will be inserted
		int tarIndex = curIndex - 1;		

		// Current element to be sorted
		int curElement = v[curIndex];

		while (tarIndex >= 0 && v[tarIndex] > curElement) {
			v[tarIndex + 1] = v[tarIndex];
			--tarIndex;
		}

		v[tarIndex + 1] = curElement;
	}
}
    \end{lstlisting}
    
    \item Implement the merge sort algorithm
    
        \begin{lstlisting}
void merge(std::vector<int>& v, int left, int mid, int right) {
	std::vector vLeft(v.begin() + left, v.begin() + mid + 1);
	std::vector vRight(v.begin() + mid + 1, v.begin() + right + 1);

	int nLeft = vLeft.size(), nRight = vRight.size();
	int vIndex = left, lIndex = 0, rIndex = 0;

	while (lIndex < nLeft && rIndex < nRight) {
		v[vIndex++] = vLeft[lIndex] < vRight[rIndex] ? vLeft[lIndex++] : vRight[rIndex++];
	}

	while (lIndex < nLeft) {
		v[vIndex++] = vLeft[lIndex++];
	}
	
	while (rIndex < nRight) {
		v[vIndex++] = vRight[rIndex++];
	}
}

void mergeSort(std::vector<int>& v, int left, int right) {
	if (left < right) {
		int mid = (right + left) / 2;

		mergeSort(v, left, mid);
		mergeSort(v, mid + 1, right);

		merge(v, left, mid, right);
	}
}

void mergeSort(std::vector<int>& v) {
	mergeSort(v, 0, v.size() - 1);
}
    \end{lstlisting}
    
    \item Just compare them? No !!! Run some simulations or tests and find the average input size for which the merge sort is an asymptotically ``better'' sorting algorithm.
\end{itemize}

Note: Include (.tex) and attach(.cpp) your source code and use a dockerfile to interact with python and plot your results.\\

\emph{BONUS:} Compare both algorithms against any other sorting algorithm

\section{Research}

Everybody at this point remembers the quadratic ``grade school'' algorithm to multiply 2 numbers of $k_{1}$ and $k_{2}$ digits respectively. \\

Your assignment now is to compare the number of operations performed by the quadratic grade school algorithm and Karatsuba multiplication.

\begin{itemize}
    \item Define Karatsuba multiplication
    \item Implement grade school multiplication
    \item Implement Karatsuba multiplication
    \item Compare Karatsuba algorithm against grade school multiplication
    \item Use any of your implemented algorithms to multiply $a*b$ where:
    \begin{description}
    \item{a:} 3141592653589793238462643383279502884197169399375105820974944592
    \item{b:} 2718281828459045235360287471352662497757247093699959574966967627
    \end{description}
\end{itemize}

Note: Include(.tex) and attach(.cpp) your source code, of course.\\

\emph{BONUS:} How about Sch\"{o}nhage-Strassen algorithm ? 

\section{Wrapping up}

Arrange the following functions in increasing order of growth rate with $g(n)$ following $f(n)$ if $f(n) = \mathcal{O}(g(n))$

\begin{enumerate}
    \item $n^{2}$
    \item $n^{2}log(n)$
    \item $n^{log(n)}$
    \item $2^{n}$
    \item $2^{2^{n}}$
\end{enumerate}

\textbf{Explanation:}

First of all, both $n^2$ and $n^2 \log(n)$ share the same factor $n^2$. If both expressions are divided by this factor, the remaining expressions would be 1 and $\log(n)$ respectively. This would suggest that $n^2 \log(n)$ has a higher growth rate than $n^2$.

Using the same approach,  $n^2 \log(n)$ and $n^{\log (n)}$ can be compared by finding common factors. For this reason, $n^{\log (n)}$ can be transformed to $n^2 n^{\log (n) - 2}$ to meet this requirement. After removing $n^2$ from the equation, we are left with $\log(n)$ from the first expression and $n^{\log (n) - 2}$ from the second one. It can be easily confirmed that

\begin{align*}
log (n) &= O(n) \\
n &= O(n^{\log (n) - 2})
\end{align*}

By the property of transitivity, we can conclude that $log (n) = O(n^{\log (n) - 2})$ which implies $log (n) = O(n^{\log (n)})$.

The comparison of functions $n^{\log (n)}$ and $2^{n}$ can be addressed with logarithmic transformations.

\begin{align*}
n^{\log (n)} &= (2 ^ {log_2(n)})^{log (n)} \\
&= 2 ^ {{log_2(n)} {log_2(n)}}\\
&= 2 ^ {log_2(n) ^ 2}
\end{align*}

As both $2 ^ {log_2(n) ^ 2}$ and $2^{n}$ share the same base, the difference in growing factors can be found in exponents $log_2(n) ^ 2$ and $n$. When replacing $n$ by $\sqrt{n}^2$, the expressions remaining (after removing the common 2 exponent) are $log_2(n)$ and $\sqrt{n}$, from which $\sqrt{n}$ is well known to have a higher growing rate than $log_2(n)$. From this result we can conclude that $n^{\log (n)} = O(2^{n})$.

Finally, it is trivial to show that $2^n = O(2^{2^n}$. The exponent of the first expression has a linear growing rate, while the exponent of the second one has an exponential growing rate. In this way, we can confirm that $2^{2^n}$ has a higher growing rate than $2^n$.

\end{document}
